---
title: "Masters Homework 2"
author: "Arturo Miranda"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}
nyc <- read.csv("NYC_HANES_DIAB.csv")
summary(nyc$MERCURYTOTALBLOOD)
#> summary(nyc$MERCURYTOTALBLOOD)
    #Min.  1st Qu.   Median     Mean  3rd Qu.     Max.     NA's 
  #0.0919   0.7564   1.6044   2.6162   3.1284 129.2344      326 

min_value <- min(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
max_value <- max(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
#> min_value
#[1] 0.0919
#> max_value
#[1] 129.2344
```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
library(e1071)
sd_mercury <- sd(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
#> sd_mercury 
#[1] 4.654974
skew_mercury <- skewness(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
#> skew_mercury 
#[1] 17.40523
```



>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
install.packages("ggplot2")
library(ggplot2)
install.packages("dplyr")
library(dplyr)
```





>- What two commands are needed to install and use a package in R?

```{r}
#install.packages()
#library()
```




>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?

```{r}
#> class(nyc)
#[1] "data.frame"
```


>- What command can you use to determine what type of data is in a column?

```{r}
#Class command: 
class(nyc$MERCURYTOTALBLOOD)
#Other option we learned:
typeof(nyc$MERCURYTOTALBLOOD)
```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(nyc$DX_DBTS)
```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}
class(nyc[["DX_DBTS"]])
```



>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}
#R will coerce every column to the same type before filling the matrix. Because the first column is character (KEY), everything else will be turned into character strings.This is because all elements must share the same type.
matrix <- as.matrix(nyc[, 1:4])
typeof(matrix)
```


>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
vec <- c(3, 9, 4, 7, 8, 9)
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)
result <- vec * WannaMatrix
result [3,3]
#> result [3,3]
#[1] 4
```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}
vec2 <- c(3,9,4,7,8)
newresult <- vec2 * WannaMatrix
newresult [3,3]
#Warning: longer object length is not a multiple of shorter object length
#> newresult [3,3]
#[1] 8

#WannaMatrix[3,3] is 1. So What’s happening: R flattens the 6×4 matrix column‑wise (24 values) and recycles the length‑5 vector (3,9,4,7,8) to match. The linear index for (3,3) is 15, so it picks the 15th element of the recycled vector, which is the 5th element of the original vec2 (8). That 8 then multiplies the matrix’s (3,3) value (1), giving 8.
```


>- Look up the arguments for read.csv

```{r}
args(read.csv)
#> args(read.csv)
#function (file, header = TRUE, sep = ",", quote = "\"", dec = ".", 
    #fill = TRUE, comment.char = "", ...) 
#NULL
```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}
median_vals <- apply(
  nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")],
  2,
  median,
  na.rm = TRUE
)

#> median_vals 
         #CADMIUM             LEAD CHOLESTEROLTOTAL        GLUCOSESI 
          #0.2936           1.0576         183.0000           4.8800 
    #CREATININESI       CREATININE 
         #70.7200           0.8000 
```


>- Do the same using sapply()

```{r}
cols <- c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", 
          "GLUCOSESI", "CREATININESI", "CREATININE")

median_vals2 <- sapply(
  nyc[ , cols],
  median,
  na.rm = TRUE
)

#> median_vals2 
         #CADMIUM             LEAD CHOLESTEROLTOTAL        GLUCOSESI 
          #0.2936           1.0576         183.0000           4.8800 
    #CREATININESI       CREATININE 
         #70.7200           0.8000 
```

>- Use a for loop to find the median value of columns 16:19

```{r}
col_idx <- 16:19

medians_16_19 <- numeric(length(col_idx))

for (k in seq_along(col_idx)) {
  medians_16_19[k] <- median(nyc[[ col_idx[k] ]], na.rm = TRUE)
}

names(medians_16_19) <- names(nyc)[col_idx]

#> medians_16_19
#CHOLESTEROLTOTAL        GLUCOSESI     CREATININESI       CREATININE 
          #183.00             4.88            70.72             0.80 


```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
cols_sd <- c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

sd_values <- numeric(length(cols_sd))

for (k in seq_along(cols_sd)) {
  sd_values[k] <- sd(nyc[[ cols_sd[k] ]], na.rm = TRUE)
}

names(sd_values) <- cols_sd
#> sd_values
#CHOLESTEROLTOTAL        GLUCOSESI     CREATININESI       CREATININE 
      #38.6872845        1.6677120       26.5131524        0.2999403 
```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}
cols_sd <- c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

sd_values <- setNames(vector("numeric", length(cols_sd)), cols_sd)

for(col in cols_sd) {
  sd_values[col] <- sd(nyc[[col]], na.rm = TRUE)
}

#> sd_values
#CHOLESTEROLTOTAL        GLUCOSESI     CREATININESI       CREATININE 
      #38.6872845        1.6677120       26.5131524        0.2999403 

```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}
nyc$A1C_status <- ifelse(nyc$A1C >= 7, "High", "Normal")
#> table(nyc$A1C_status, useNA = "ifany")

  #High Normal   <NA> 
    #79   1125    323 
```
###################################################################################OPTIONAL##############################################################################################

>- Subset to only have females (gender == "female")

```{r}

```


>- Create a subset of males and find the median total cholesterol level

```{r}

```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

```

Find the class of the first three columns of data. What are they?

```{r}

```

